\Huge \bf \sc Design and implementation of sensor data fusion for an autonomous quadrotor

### 2013-11-21 latex/paper_sensor_fusion_i2mtc/paper_sensor_fusion_i2mtc.tex
papers actualizados

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{\Huge \bf \sc
Design and implementation of sensor fusion and control for an autonomous quadrotor}

@@

%\vspace*{-.5cm}
\begin{abstract}
%\boldmath
This paper describes the design and integration of an instrumentation, sensor fusion and control system that allows the autonomous flight of a quadrotor. A comercial frame is used, a mathematical model for the quadrotor is developed and its parameters determined from the characterization of the unit. An intelligence is integrated to play the role of the flight controller. A 9 degrees of freedom Inertial Measurement Unit (IMU) equipped with a barometer is calibrated and added to the platform. Sensor fusion is done by two modified Extended Kalman Filters (EKF): one combining data provided by IMU and the other also including the information provided by GPS. A reliable estimation of the state variables is obtained. Three states representing systematic bias in the accelerometer measurements are also added to the EKF, which improves the inertial estimation of the position. The control actions are obtained from a proportional-integral controller based on an improved Linear Quadratic Regulator (LQR) algorithm. A stable autonomous platform is achieved.%For windless environments, a stable platform is achieved.
% -cambio-
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

The purpose of the reported work is to develop the instrumentation and control of an autonomous quadrotor in which is possible to modify and test different state estimation techniques, path planning algorithms and control laws. The solution was reached using a commercial mechanical platform, and design and build an ad-hoc instrumentation and control navigation system.
An aerial vehicle is inherently unstable, staying still is not a simple task. This paper focuses on the stabilization of a quadrotor, explaining the development of the mathematical model to represent the system, the filtering techniques applied for sensor data fusion and the control system that allows the quadrotor to fly.

@@

The dynamic system follows the model:
\begin{eqnarray}
	\mathbf{x}_{k}^e &=& f(\mathbf{x}_{k-1}^e, \mathbf{u}_{k-1}) + \boldsymbol{\eta}^w_{k-1}\\
	\mathbf{z}_{k} &=& h(\mathbf{x}_{k}) + \boldsymbol{\eta}^{v}_{k}
\end{eqnarray}
where $\mathbf{z}_{k}$ is the observation at time $k$. The function $f$, based on the dynamics of the system (equations \refp{eq:modelo}), is deduced form equation \refp{ec:fxu} and is used to compute the predicted state from the previous one. The function $h$ is used to compute the predicted measurement from the predicted state. In other words, $f$ keeps the information about state evolution, and $h$ represents the transformation between the state vector and the ideal (noiseless) observation. The state transition and observation models do not need to be linear functions, but must be differentiable. $f$ and $h$ cannot be directly applied to the covariance. Instead a partial differential matrix (the \emph{Jacobian}) is computed, which is evaluated in the current state at each time step. Due to dynamical linearization, the EKF cannot ensure statistic optimality, because the precision highly depends on the linearization precision \cite{bib:kay}.

@@

\subsubsection{Prediction and update equations for $EKF_{IMU}$}
\begin{itemize}
	\item Prediction:
	\begin{itemize}
		\item $\hat{\mathbf{x}}_{k|k-1}^e = f(\hat{\mathbf{x}}_{k-1|k-1}^e, \mathbf{u}_{k-1})$
		\item $ \mathbf{P}_{k|k-1} =  {{\mathbf{F}_{k-1}}} \mathbf{P}_{k-1|k-1}{  {\mathbf{F}_{k-1}^T}} + \mathbf{Q}_{k-1} $
	\end{itemize}
	\item Update
	\begin{itemize}
		\item $\tilde{\mathbf{y}}_{k} = \mathbf{z}_{k} - h(\hat{\mathbf{x}}_{k|k-1}^e)$
		\item $\mathbf{S}_{k} = { \mathbf{H}_{k}}\mathbf{P}_{k|k-1}{ \mathbf{H}_{k}^\top} + \mathbf{R}_{k}$
		\item $\mathbf{K}_{k} = \mathbf{P}_{k|k-1}{ \mathbf{H}_{k}^\top}\mathbf{S}_{k}^{-1} $
		\item $\hat{\mathbf{x}}_{k|k}^e = \hat{\mathbf{x}}_{k|k-1}^e + \mathbf{K}_{k}\tilde{\mathbf{y}}_{k} $
		\item $ \mathbf{P}_{k|k} = (I - \mathbf{K}_{k} { \mathbf{H}_{k}}) \mathbf{P}_{k|k-1} $
	\end{itemize}
\end{itemize}
where $\mathbf{z}_{k}$ is the observation at time $k$.\\
The state and observation transition matrices are 
$$ 
{ \mathbf{F}_{k-1}} = \left . \frac{\partial f}{\partial \mathbf{x}^e } \right \vert _{\hat{\mathbf{x}}_{k-1|k-1}^e,\mathbf{u}_{k-1}} , \;\;\;\;\;\;\;
{ \mathbf{H}_{k}} = \left . \frac{\partial h}{\partial \mathbf{x}^e } \right \vert _{\hat{\mathbf{x}}_{k|k-1}^e} 
$$
\subsubsection{Prediction and update equations for $EKF_{IMU+GPS}$}
\begin{itemize}
	\item Prediction:
	\begin{itemize}
		\item $\hat{\mathbf{x}}_{k|k-1}^e = f_{G}(\hat{\mathbf{x}}_{k-1|k-1}^e, \mathbf{u}_{k-1})$
		\item $ \mathbf{P_{G}}_{k|k-1} =  {{\mathbf{F_{G}}_{k-1}}} \mathbf{P_{G}}_{k-1|k-1}{  {\mathbf{F_{G}}_{k-1}^T}} + \mathbf{Q_{G}}_{k-1} $
	\end{itemize}
	\item Update
	\begin{itemize}
		\item $\mathbf{\tilde{y_G}}_{k} = \mathbf{z_{G}}_{k} - h_G(\hat{\mathbf{x}}_{k|k-1}^e)$
		\item $\mathbf{S_{G}}_{k} = { \mathbf{H_{G}}_{k}}\mathbf{P_{G}}_{k|k-1}{ \mathbf{H_{G}}_{k}^\top} + \mathbf{R_{G}}_{k}$
		\item $\mathbf{K_{G}}_{k} = \mathbf{P_{G}}_{k|k-1}{ \mathbf{H_{G}}_{k}^\top}\mathbf{S_{G}}_{k}^{-1} $
		\item $\hat{\mathbf{x}}_{k|k}^e = \hat{\mathbf{x}}_{k|k-1}^e + \mathbf{K_{G}}_{k}\tilde{\mathbf{y}}_{k} $
		\item $ \mathbf{P_{G}}_{k|k} = (I - \mathbf{K_{G}}_{k} { \mathbf{H_{G}}_{k}}) \mathbf{P}_{k|k-1} $
	\end{itemize}
\end{itemize}
The two sets of prediction and update equations, for $EKF_{IMU}$ and $EKF_{IMU+GPS}$ are very similar, and the subindex $_G$ is used to explicitly show that the variables are different in each filter. This difference is caused because $\mathbf{z}_{k}$ and $\mathbf{z_G}_{k}$ are different since $\mathbf{z_G}_{k}$ includes GPS data. Note that $\hat{\mathbf{x}}_{k|k-1}^e$ is always the same.

@@

\subsection{Results}
%\label{sec:kalman-ekf}

@@

In each graphic of Figure \refp{fig:kalman} is shown the primary estimation obtained from sensors and the kalman estimation of that state variable for the three Euler angles. As can be seen, the noise is greatly reduced and no delay is introduced.\\

@@

\section{Control design}
\label{sec:control}

To work directly with equations \refp{eq:modelo}, a non-linear control technique would be required. To simplifly the control system, equations \refp{eq:modelo} are linearized near certain points of operation which result in a Linear Time Invariant (LTI) system of the form:
\begin{equation}
  \label{eq:lti}
  \mathbf{\dot{x}} = A\mathbf{x} + B\mathbf{u}
\end{equation}

This restricts the system to: Hovering\footnote{Constant state vector.}, uniform linear trajectories, and uniform circular trajectories around the vertical axis. Working with an LTI greatly simplifies the design of the control system and does it not introduce significant limitations.

The control system is shown in Figure \refp{fig:diagrama_bloques_eng.pdf}. The feedback matrices have many entries, $K_p$ has 48 and $K_i$ has 16. A \textit{root locus} or \textit{pole-zero} analysis on such a system is not a simple task, additional complications are introduced by the fact that $K_p$ and $K_i$ will change when the trajectory changes, they depend on the linearization of the system. A simple, automated and very commun solution is achieved by means of the LQR algorithm (see e.g.: \cite{bib:lqr-discreto}). In this work an extended state vector that includes the integrals of some state variables was introduce to take into consideration errors in the characterization and modeling of the system.

\begin{figure}
%	\vspace{-60pt}
	\centering
	\includegraphics[width=1\columnwidth]{./pics_paper/diagrama_bloques_eng.png}
%	\vspace{-180pt}
	\caption{\textbf{Control System:} $X_{sp}$ and $U_{sp}$ represent the set-points for the state vector and the speed of each motor. The $K_p$ and $K_i$ blocks are the proportional and integral gain matrices. The output of the system is the current state vector.}
	\label{fig:diagrama_bloques_eng.pdf}
	\vspace{-0.5cm}
\end{figure}

\subsection{The LQR algorithm}
\label{sec:lqr}

A a system which is controllable and observable is guaranteed to be stable and for this type of system, the LQR algorithm provides an optimal controller (see e.g.: \cite{bib:lqrnotes}) by minimizing the following cost function:
\begin{equation}
\label{eq:lqr}
\int_{0}^{\infty}  (\mathbf{x}-\mathbf{x}_{sp})^T\mathbf{Q} (\mathbf{x}-\mathbf{x}_{sp})+(\mathbf{u}-\mathbf{u}_{sp})^T\mathbf{R}(\mathbf{u}-\mathbf{u}_{sp})dt
\end{equation}

where $\mathbf{Q}$ and $\mathbf{R}$ are positive-definite matrices, weighting the energy of the departure from the objective state and the objective control signal respectively. The choice of $\mathbf{Q}$ and $\mathbf{R}$ is a trade-off between the error that is tolerated and the energy that the system can use to correct it. In equation \refp{eq:lqr} $\mathbf{x}_{sp}$ is the objective state and $\mathbf{u}_{sp}$ is the theoretical control action that is compatible with the state $\mathbf{x}_{sp}$. Both vectors as well as $\mathbf{x}$ and $\mathbf{u}$ are time dependent.

 When the system is a feedback system with known state variables, like the one shown in Figure \refp{fig:diagrama_bloques_eng.pdf}, the solution is given by \refp{eq:lqr-sol}:

\begin{equation}
\begin{array}{c}
\mathbf{u}(t)-\mathbf{u}_{sp}(t) = -K(\mathbf{x}(t)-\mathbf{x}_{sp}(t))\\
\\
\mathbf{K} = \mathbf{R}^{-1}\mathbf{B}^T\mathbf{P}
\end{array}
\label{eq:lqr-sol}
\end{equation}

where $\mathbf{P}$ is the solution to Riccati's equation:
\begin{equation}
  \label{eq:riccati}
  \mathbf{A}^T\mathbf{P} + \mathbf{P}\mathbf{A} - \mathbf{P}\mathbf{B}\mathbf{R}^{-1}\mathbf{B}^T\mathbf{P} + \mathbf{Q} = 0
\end{equation}

\subsection{Integral effect}
In this work an integral effect is added over the LQR controller. For this, an expanded state vector is defined $[\mathbf{x}^T, \mathbf{x}_I^T]^T$. For the controllability and observability hypothesis to hold, only some state variables may be passed through the integral control block: $\mathbf{\dot{x}}_I =[x,y,z,\theta]^T$. After making this definitions, the system is re-written in state variables form. The method to obtain the new matrix is the same that explained in Section \refp{sec:lqr}
\section{Software}

\subsection{Architecture}
\label{sec:software-arch}

The software runs on several independent boards:

\begin{itemize}
\item \textbf{Flight controller:} Run by an ARM-Cortex-A8. Processes the data from the IMU, calculates the necessary control actions, and sets the desired speed for each motor by sending the appropriate commands, via $I^2C$, to the ESCs. It requires Linux specific \textit{C} functions to handle I/O. The design of the flight controller is modular, thus replacing any part of it should be a straightforward task. The most critical aspect of the flight controller is timing, any delays will severely degrade the performance of the system.
\item \textbf{IMU:} Uses an ATMega328p on the IMU. The firmware on the IMU takes care of reading from all the sensors (using $I^2C$) and sending a new frame of data every 10ms over a UART.
\item \textbf{ESCs:} Four microprocessors\footnote{C8051F330/1 - Silicon Labs.}, one for each motor.
\end{itemize}

The flight controller is accesed via an \textit{ssh} session over \textit{WiFi}. Once logged in, the flight controller software can be compiled and executed.

\subsection{Implementation details}
\label{sec:software-impl}

The code implements a discrete version of the LQR control system mentioned in \refp{sec:control}, based on a discretization of the model described in \refp{sec:modelo}. There are some practical considerations worth mentioning:
\begin{itemize}
\item \textbf{Anti-wind-up:} In order to avoid divergency of integral terms an Anti-wind-up algorithm was implemented.
\item \textbf{Limiting:} The control block sets $\vert \vec{X} - \vec{X}_{sp}\vert < \vert \vec{X}_{th} \vert$ to avoid attempts to perform actions that the system cannot handle.
\end{itemize}

\section{Results}
\label{sec:results-tests}

\subsection{Basic stability}

During this first stage, the quadrotor was only allowed to move (and correct) one of the following Euler angles: $\{\psi,\phi\}$. The control system was tuned until an acceptable behavior was obtained. This procedure was applied to each angle. Once this stage was completed the quadrotor was able to maintain horizontality by controlling $\{\psi,\phi,\omega_{q_x},\omega_{q_y}\}$. The behavior for the Roll ($\psi$) angle is shown in Figure \refp{fig:psimetro}. While in Figure \refp{fig:psimetro_psi} is shown the mechanical perturbation imposed to the \emph{Roll} angle, the reaction can be observed in Figure \refp{fig:psimetro_w}. From $t_0 = 80.9 s$ to $t_1 =81.3 s$ the perturbation is present. After this, the controller is working freely. The system step response for a step of more than $20^o$ shows a overshoot of $3^o$ or $4^o$ and a rise time of about $0.4\;s$, which means an excellent performance, as can be seen in Figure \refp{fig:psimetro_psi}.
\begin{figure}[h!]
  \centering
  \subfloat[Roll angle deviation]{\label{fig:psimetro_psi}
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/psimetro_psi.pdf}} \hspace*{-5pt}
  \subfloat[Difference between $\omega$ and $\omega_{hover}$ for each motor]{\label{fig:psimetro_w} 
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/psimetro_w.pdf}}
  \caption{\textbf{Stablization experiment -} The Roll angle is deviated from equilibrium and the quadrotor manage to stabilize itself.}
  \label{fig:psimetro}
\end{figure}

\subsection{Orientation}
To verify orientation was correctly maintained, the speed of the motors was reduced below the level required to lift off, and the quadrotor was hung from strings. After this test, the control system was able to hold the last of the Euler angles: $\{\theta\}$. 
\begin{figure}[h!]
  \centering
  \subfloat[Yaw angle deviation]{\label{fig:theta}
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/theta.pdf}} \hspace*{-5pt}
  \subfloat[Difference between $\omega$ and $\omega_{hover}$ for each pair of motors.]{\label{fig:theta_w} 
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/theta_w.pdf}}
  \caption{\textbf{Orientation experiment -} The Yaw angle is deviated from equilibrium and the quadrotor manage to stabilize itself.}
  \label{fig:theta_all}
\end{figure}
In the Figure \refp{fig:theta_all} is shown the system behavior against a mechanical perturbation (from $t_0 = 59.3 s$ to $t_1 = 60.5 s$) in yaw angle. To rotate, the system must generate a torque in the opposite direction of the perturbation by unbalancing the angular velocity of the motors in pairs: $\omega_1$ and $\omega_3$ in one direction and $\omega_2$ and $\omega_4$ in the other. In this case, the rise time is of about $1.6\;s$, much higher than in the other two Euler angles, which can be permitted as Yaw angle is not such critical for stability. The overshoot obtained is of about 13\% of the step size.

\subsection{Limited Flying}
\label{sec:results-system-response}
At this point the three Euler angles are controlled, so a basic level of stability is expected, and the quadrotor is allowed to fly. After this test, the control system was able to take-off and to hold altitud and vertical speed: $\{z,v_{q_z}\}$.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=.7\columnwidth]{./pics_paper/altura.pdf}
	\caption{\textbf{Altitud:} Performance during takeoff (from 0m to 1m).}
	\label{fig:altura.pdf}
\end{figure}

@@

\section{Conclusion}
\balance
The final plataform implemented is shown in Figure \refp{fig:tito}. An acrylic box was added in the bottom of the quadrotor to hold the added intelligence. The GPS is attached at the upper part of the quadrotor, so as it has the best satellite visibility. Aluminum protections were added around all motors and propellers so as damages can be prevented.\\

The two main goals were successfuly achieved. First, the design and implementation of the sensor fusion was accomplished by the integration of all the measurements in a modified Extended Kalman Filter which is used for denoising purposes without introducing delay. In addition a basic controller, capable of stabilizing the system, was implemented based on clasic LQR algorithm and adding extra states to perform also an integral control action. A simplified dynamic model was derived as a first step for both Kalman Filter and controller. \\

The Extended Kalman Filter implemented gave very good results for all the state variables except for the linear velocities on axes $\{i_q,\;j_q\}$. The main reason is that a velocity sensor is not available, thus, it can only be infered by integrating the accelerations. Therefore, an extra sensor should be used (i.e. a more accurate GPS) or a more sophisticated algorithm should be develped. \\

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{\Huge \bf \sc
Design and implementation of sensor data fusion for an autonomous quadrotor}

@@

%\vspace*{-.5cm}
\begin{abstract}
%\boldmath
This paper describes the design and integration of an instrumentation and sensor fusion that is used to allow the autonomous flight of a quadrotor. A comercial frame is used, a mathematical model for the quadrotor is developed and its parameters determined from the characterization of the unit. A 9 degrees of freedom Inertial Measurement Unit (IMU) equipped with a barometer is calibrated and added to the platform. Sensor fusion is done by two modified Extended Kalman Filters (EKF): one combining data provided by IMU and the other also including the information provided by GPS. A reliable estimation of the state variables is obtained. Three states representing systematic bias in the accelerometer measurements are also added to the EKF, which improves the inertial estimation of the position. A stable autonomous platform is achieved.%For windless environments, a stable platform is achieved.
% -cambio-
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

This work is part of the development of the instrumentation and control of an autonomous quadrotor in which is possible to modify and test different state estimation techniques, path planning algorithms and control laws. The solution was reached using a commercial mechanical platform, and design and build an ad-hoc instrumentation and control navigation system.
This paper focuses on the filtering techniques applied for sensor data fusion which gives the state estimation needed by the control system, that allows the quadrotor to fly.

@@

The dynamic system follows the model:
\begin{eqnarray}
	\mathbf{x}_{k}^e &=& f(\mathbf{x}_{k-1}^e, \mathbf{u}_{k-1}) + \boldsymbol{\eta}^w_{k-1}\\
	\mathbf{z}_{k} &=& h(\mathbf{x}_{k}) + \boldsymbol{\eta}^{v}_{k}
\end{eqnarray}
where $\mathbf{z}_{k}$ is the observation at time $k$. The function $f$, based on the dynamics of the system is deduced form equation \refp{ec:fxu} and is used to compute the predicted state from the previous one. The function $h$ is used to compute the predicted measurement from the predicted state. In other words, $f$ keeps the information about state evolution, and $h$ represents the transformation between the state vector and the ideal (noiseless) observation. The state transition and observation models do not need to be linear functions, but must be differentiable. $f$ and $h$ cannot be directly applied to the covariance. Instead a partial differential matrix (the \emph{Jacobian}) is computed, which is evaluated in the current state at each time step. Due to dynamical linearization, the EKF cannot ensure statistic optimality, because the precision highly depends on the linearization precision \cite{bib:kay}.

@@

\subsubsection{Prediction and update equations for $EKF_{IMU}$}
\begin{itemize}
	\item Prediction:
	\begin{itemize}
		\item $\hat{\mathbf{x}}_{k|k-1}^e = f(\hat{\mathbf{x}}_{k-1|k-1}^e, \mathbf{u}_{k-1})$
		\item $ \mathbf{P}_{k|k-1} =  {{\mathbf{F}_{k-1}}} \mathbf{P}_{k-1|k-1}{  {\mathbf{F}_{k-1}^T}} + \mathbf{Q}_{k-1} $
	\end{itemize}
	\item Update
	\begin{itemize}
		\item $\tilde{\mathbf{y}}_{k} = \mathbf{z}_{k} - h(\hat{\mathbf{x}}_{k|k-1}^e)$
		\item $\mathbf{S}_{k} = { \mathbf{H}_{k}}\mathbf{P}_{k|k-1}{ \mathbf{H}_{k}^\top} + \mathbf{R}_{k}$
		\item $\mathbf{K}_{k} = \mathbf{P}_{k|k-1}{ \mathbf{H}_{k}^\top}\mathbf{S}_{k}^{-1} $
		\item $\hat{\mathbf{x}}_{k|k}^e = \hat{\mathbf{x}}_{k|k-1}^e + \mathbf{K}_{k}\tilde{\mathbf{y}}_{k} $
		\item $ \mathbf{P}_{k|k} = (I - \mathbf{K}_{k} { \mathbf{H}_{k}}) \mathbf{P}_{k|k-1} $
	\end{itemize}
\end{itemize}
where $\mathbf{z}_{k}$ is the observation at time $k$.\\
The state and observation transition matrices are 
$$ 
{ \mathbf{F}_{k-1}} = \left . \frac{\partial f}{\partial \mathbf{x}^e } \right \vert _{\hat{\mathbf{x}}_{k-1|k-1}^e,\mathbf{u}_{k-1}} , \;\;\;\;\;\;\;
{ \mathbf{H}_{k}} = \left . \frac{\partial h}{\partial \mathbf{x}^e } \right \vert _{\hat{\mathbf{x}}_{k|k-1}^e} 
$$
\subsubsection{Prediction and update equations for $EKF_{IMU+GPS}$}
\begin{itemize}
	\item Prediction:
	\begin{itemize}
		\item $\hat{\mathbf{x}}_{k|k-1}^e = f_{G}(\hat{\mathbf{x}}_{k-1|k-1}^e, \mathbf{u}_{k-1})$
		\item $ \mathbf{P}_{k|k-1} =  {{\mathbf{F_{G}}_{k-1}}} \mathbf{P}_{k-1|k-1}{  {\mathbf{F_{G}}_{k-1}^T}} + \mathbf{Q}_{k-1} $
	\end{itemize}
	\item Update
	\begin{itemize}
		\item $\mathbf{\tilde{y_G}}_{k} = \mathbf{z_{G}}_{k} - h_G(\hat{\mathbf{x}}_{k|k-1}^e)$
		\item $\mathbf{S_{G}}_{k} = { \mathbf{H_{G}}_{k}}\mathbf{P_{G}}_{k|k-1}{ \mathbf{H_{G}}_{k}^\top} + \mathbf{R_{G}}_{k}$
		\item $\mathbf{K_{G}}_{k} = \mathbf{P_{G}}_{k|k-1}{ \mathbf{H_{G}}_{k}^\top}\mathbf{S_{G}}_{k}^{-1} $
		\item $\hat{\mathbf{x}}_{k|k}^e = \hat{\mathbf{x}}_{k|k-1}^e + \mathbf{K_{G}}_{k}\tilde{\mathbf{y}}_{k} $
		\item $ \mathbf{P}_{k|k} = (I - \mathbf{K_{G}}_{k} { \mathbf{H_{G}}_{k}}) \mathbf{P}_{k|k-1} $
	\end{itemize}
\end{itemize}
The two sets of prediction and update equations, for $EKF_{IMU}$ and $EKF_{IMU+GPS}$ are very similar, and the subindex $_G$ is used to explicitly show that the variables are different in each filter. This difference is caused because $\mathbf{z}_{k}$ and $\mathbf{z_G}_{k}$ are different since $\mathbf{z_G}_{k}$ includes GPS data. Note that $\hat{\mathbf{x}}_{k|k-1}^e$ is always the same.

@@

\section{Results}

\subsection{Sensor fusion}
%\label{sec:kalman-ekf}

@@

In each graphic of Figure \refp{fig:kalman} is shown the primary estimation obtained from sensors and the kalman estimation of that state variable for the three Euler angles. As can be seen, the noise is greatly reduced and no delay is introduced.

@@

% \section{Control design}
% \label{sec:control}
% 
% To work directly with equations \refp{eq:modelo}, a non-linear control technique would be required. To simplifly the control system, equations \refp{eq:modelo} are linearized near certain points of operation which result in a Linear Time Invariant (LTI) system of the form:
% \begin{equation}
%   \label{eq:lti}
%   \mathbf{\dot{x}} = A\mathbf{x} + B\mathbf{u}
% \end{equation}
% 
% This restricts the system to: Hovering\footnote{Constant state vector.}, uniform linear trajectories, and uniform circular trajectories around the vertical axis. Working with an LTI greatly simplifies the design of the control system and does it not introduce significant limitations.
% 
% The control system is shown in Figure \refp{fig:diagrama_bloques_eng.pdf}. The feedback matrices have many entries, $K_p$ has 48 and $K_i$ has 16. A \textit{root locus} or \textit{pole-zero} analysis on such a system is not a simple task, additional complications are introduced by the fact that $K_p$ and $K_i$ will change when the trajectory changes, they depend on the linearization of the system. A simple, automated and very commun solution is achieved by means of the LQR algorithm (see e.g.: \cite{bib:lqr-discreto}). In this work an extended state vector that includes the integrals of some state variables was introduce to take into consideration errors in the characterization and modeling of the system.
% 
% \begin{figure}
% %	\vspace{-60pt}
% 	\centering
% 	\includegraphics[width=1\columnwidth]{./pics_paper/diagrama_bloques_eng.png}
% %	\vspace{-180pt}
% 	\caption{\textbf{Control System:} $X_{sp}$ and $U_{sp}$ represent the set-points for the state vector and the speed of each motor. The $K_p$ and $K_i$ blocks are the proportional and integral gain matrices. The output of the system is the current state vector.}
% 	\label{fig:diagrama_bloques_eng.pdf}
% 	\vspace{-0.5cm}
% \end{figure}
% 
% \subsection{The LQR algorithm}
% \label{sec:lqr}
% 
% A a system which is controllable and observable is guaranteed to be stable and for this type of system, the LQR algorithm provides an optimal controller (see e.g.: \cite{bib:lqrnotes}) by minimizing the following cost function:
% \begin{equation}
% \label{eq:lqr}
% \int_{0}^{\infty}  (\mathbf{x}-\mathbf{x}_{sp})^T\mathbf{Q} (\mathbf{x}-\mathbf{x}_{sp})+(\mathbf{u}-\mathbf{u}_{sp})^T\mathbf{R}(\mathbf{u}-\mathbf{u}_{sp})dt
% \end{equation}
% 
% where $\mathbf{Q}$ and $\mathbf{R}$ are positive-definite matrices, weighting the energy of the departure from the objective state and the objective control signal respectively. The choice of $\mathbf{Q}$ and $\mathbf{R}$ is a trade-off between the error that is tolerated and the energy that the system can use to correct it. In equation \refp{eq:lqr} $\mathbf{x}_{sp}$ is the objective state and $\mathbf{u}_{sp}$ is the theoretical control action that is compatible with the state $\mathbf{x}_{sp}$. Both vectors as well as $\mathbf{x}$ and $\mathbf{u}$ are time dependent. 
% 
%  When the system is a feedback system with known state variables, like the one shown in Figure \refp{fig:diagrama_bloques_eng.pdf}, the solution is given by \refp{eq:lqr-sol}:
% 
% \begin{equation}
% \begin{array}{c}
% \mathbf{u}(t)-\mathbf{u}_{sp}(t) = -K(\mathbf{x}(t)-\mathbf{x}_{sp}(t))\\
% \\
% \mathbf{K} = \mathbf{R}^{-1}\mathbf{B}^T\mathbf{P}
% \end{array}
% \label{eq:lqr-sol}
% \end{equation}
% 
% where $\mathbf{P}$ is the solution to Riccati's equation:
% \begin{equation}
%   \label{eq:riccati}
%   \mathbf{A}^T\mathbf{P} + \mathbf{P}\mathbf{A} - \mathbf{P}\mathbf{B}\mathbf{R}^{-1}\mathbf{B}^T\mathbf{P} + \mathbf{Q} = 0
% \end{equation}
% 
% \subsection{Integral effect}
% In this work an integral effect is added over the LQR controller. For this, an expanded state vector is defined $[\mathbf{x}^T, \mathbf{x}_I^T]^T$. For the controllability and observability hypothesis to hold, only some state variables may be passed through the integral control block: $\mathbf{\dot{x}}_I =[x,y,z,\theta]^T$. After making this definitions, the system is re-written in state variables form. The method to obtain the new matrix is the same that explained in Section \refp{sec:lqr}
% \section{Software}
% 
% \subsection{Architecture}
% \label{sec:software-arch}
% 
% The software runs on several independent boards:
% 
% \begin{itemize}
% \item \textbf{Flight controller:} Run by an ARM-Cortex-A8. Processes the data from the IMU, calculates the necessary control actions, and sets the desired speed for each motor by sending the appropriate commands, via $I^2C$, to the ESCs. It requires Linux specific \textit{C} functions to handle I/O. The design of the flight controller is modular, thus replacing any part of it should be a straightforward task. The most critical aspect of the flight controller is timing, any delays will severely degrade the performance of the system.
% \item \textbf{IMU:} Uses an ATMega328p on the IMU. The firmware on the IMU takes care of reading from all the sensors (using $I^2C$) and sending a new frame of data every 10ms over a UART.
% \item \textbf{ESCs:} Four microprocessors\footnote{C8051F330/1 - Silicon Labs.}, one for each motor.
% \end{itemize}
% 
% The flight controller is accesed via an \textit{ssh} session over \textit{WiFi}. Once logged in, the flight controller software can be compiled and executed.
% 
% \subsection{Implementation details}
% \label{sec:software-impl}
% 
% The code implements a discrete version of the LQR control system mentioned in \refp{sec:control}, based on a discretization of the model described in \refp{sec:modelo}. There are some practical considerations worth mentioning:
% \begin{itemize}
% \item \textbf{Anti-wind-up:} In order to avoid divergency of integral terms an Anti-wind-up algorithm was implemented.
% \item \textbf{Limiting:} The control block sets $\vert \vec{X} - \vec{X}_{sp}\vert < \vert \vec{X}_{th} \vert$ to avoid attempts to perform actions that the system cannot handle.
% \end{itemize}

% \section{Results}
\label{sec:results-tests}

\subsection{Closed loop stability: Attitude control}
The described state estimation was used by the control system, as shown in figure \refp{fig:diagrama_tio}. The control algorithm uses the state estimation as input and produces the instantaneous speed of each motor, thus establish the thrust and momentum.
The quadrotor was able to maintain horizontality by controlling $\{\psi,\phi,\omega_{q_x},\omega_{q_y}\}$. 
Figure \refp{fig:psimetro} shows the response of the quadrotor to a perturbation in the Roll ($\psi$) angle. While in Figure \refp{fig:psimetro_psi} is shown the mechanical perturbation imposed, the reaction (controller commands) can be observed in Figure \refp{fig:psimetro_w}. From $t_0 = 80.9 s$ to $t_1 =81.3 s$ the perturbation is present. After this, the controller is working freely. The system step response for a step of more than $20^o$ shows a overshoot of $3^o$ or $4^o$ and a rise time of about $0.4\;s$, which means an excellent performance, as can be seen in Figure \refp{fig:psimetro_psi}.
\begin{figure}[h!]
  \centering
  \subfloat[Roll angle deviation]{\label{fig:psimetro_psi}
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/psimetro_psi.pdf}} \hspace*{-5pt}
  \subfloat[Difference between $\omega$ and $\omega_{hover}$ for each motor]{\label{fig:psimetro_w} 
  		\includegraphics[width=0.5\columnwidth]
  			{./pics_paper/psimetro_w.pdf}}
  \caption{\textbf{Stablization experiment -} The Roll angle is deviated from equilibrium and the quadrotor manage to stabilize itself.}
  \label{fig:psimetro}
\end{figure}

% \subsection{Orientation}
% To verify orientation was correctly maintained, the speed of the motors was reduced below the level required to lift off, and the quadrotor was hung from strings. After this test, the control system was able to hold the last of the Euler angles: $\{\theta\}$. 
% \begin{figure}[h!]
%   \centering
%   \subfloat[Yaw angle deviation]{\label{fig:theta}
%   		\includegraphics[width=0.5\columnwidth]
%   			{./pics_paper/theta.pdf}} \hspace*{-5pt}
%   \subfloat[Difference between $\omega$ and $\omega_{hover}$ for each pair of motors.]{\label{fig:theta_w} 
%   		\includegraphics[width=0.5\columnwidth]
%   			{./pics_paper/theta_w.pdf}}
%   \caption{\textbf{Orientation experiment -} The Yaw angle is deviated from equilibrium and the quadrotor manage to stabilize itself.}
%   \label{fig:theta_all}
% \end{figure}
% In the Figure \refp{fig:theta_all} is shown the system behavior against a mechanical perturbation (from $t_0 = 59.3 s$ to $t_1 = 60.5 s$) in yaw angle. To rotate, the system must generate a torque in the opposite direction of the perturbation by unbalancing the angular velocity of the motors in pairs: $\omega_1$ and $\omega_3$ in one direction and $\omega_2$ and $\omega_4$ in the other. In this case, the rise time is of about $1.6\;s$, much higher than in the other two Euler angles, which can be permitted as Yaw angle is not such critical for stability. The overshoot obtained is of about 13\% of the step size.

\subsection{Closed loop stability: Limited Flying}
\label{sec:results-system-response}
At this point the three Euler angles are controlled, so a basic level of stability is expected, and the quadrotor is allowed to fly. After this test, the control system was able to take-off and to hold altitud and vertical speed: $\{z,v_{q_z}\}$.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=.7\columnwidth]{./pics_paper/altura.pdf}
	\caption{\textbf{Altitude:} Performance during takeoff (from 0m to 1m).}
	\label{fig:altura.pdf}
\end{figure}

@@

\section{Conclusion}
\balance

The main goal was successfuly achieved. The design and implementation of the sensor fusion for a quadrotor was accomplished by the integration of all the measurements in a double modified Extended Kalman Filter which is used for denoising purposes without introducing delay. A simplified dynamic model was derived as a first step for the Kalman Filter. \\

Through the implementation of the double Kalman filter, two important problems were solved: the difference between the IMU and GPS sampling rates and the effect of the bias drift in the accelerometer measurements.\\

The developed sensor fusion technique was implemented as part of a closed loop control system. It was successfully tested for attitude and height. \\

This system is an experimental platform for future research.
